---
title: "GAM air pollution"
author: "Afiqah"
date: "2023-04-07"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Work flow

## 1. Preparation
### 1.1. Load package
### 1.2. Load data

## 2. Data exploration
### 2.1. Data wrangling
#### 2.1.1. Read and describe data
#### 2.1.2. Imputation of missing data
#### 2.1.3. Change variable format
#### 2.1.4. Visualize air pollutant data
#### 2.1.5. Visualize weather data
#### 2.1.6. Total respiratory diagnosis by city
#### 2.1.7. Ranking of respiratory diseases by period
#### 2.1.8. Air pollution by city, by period
#### 2.1.9. Respiratory diseases by city, by period

### 2.2. Variable selection

#### 2.2.1. Correlation
#### 2.2.2. Comparison of air pollution and respiratory diseases between periods
#### 2.2.3. Univariate analysis

## 3. Association analysis (Generalized additive model)
### 3.1. Model selection and checking for assumptions
### 3.2. Kota Bharu analysis
### 3.3. Johor Bahru analysis

# Analysis

## 1. Preparation

### 1.1. Load package

```{r}
library(mice)
library(dplyr)
library(ggplot2)
library(ggpubr)
library(lubridate)
library(gtsummary)
library(corrplot)
library(mgcv)
library(ISLR)
library(voxel)
library(gridExtra)
library(mvmeta)
library(splines)
library(plot3D)
library(tidyverse)
library(rstatix)
library(foreign)
library(nnet)
library(AER)
library(ggbump)
library(moments)
library(MASS)
```

### 1.2. Load data

```{r}
data <- read.csv("citydata.csv")
```
## 2. Data exploration

## 2. Data exploration

### 2.1. Data wrangling

#### 2.1.1. Read and describe data

```{r}
glimpse(data)
summary(data)
```
#### 2.1.2. Imputation of missing data

```{r}
# Missing data pattern
md.pattern(data)

# Imputation of missing data
tempData <- mice(data,m=5,maxit=50,meth='pmm',seed=500)
summary(tempData)

# Explore imputated data
densityplot(tempData)

#New dataframe
data2 <- complete(tempData,1)

summary(data2)
```

#### 2.1.3 Change variable format

```{r}
data2$Date <- dmy(data2$Date)

data2$Period <- as.factor(data2$Period)
data2$City <- as.factor(data2$City)

data2$Infection <- data2$A00.B99.Certain.infectious.and.parasitic.diseases
data2$URTI <- data2$J00.J06_Acute.upper.respiratory.infections
data2$Pneumonia <- data2$J09.J18_Influenza.and.pneumonia
data2$LRTI <- data2$J20.J22_Other.acute.lower.respiratory.infections
data2$CLD <- data2$J40.J47_Chronic.lower.respiratory.diseases
data2$Pleura <- data2$J90.J94.Other.diseases.of.the.pleura
data2$Symptoms_Signs <- data2$R00.R09.Symptoms.and.signs.involving.the.circulatory.and.respiratory.systems
data2$COVID <- data2$U07.1_COVID19

kbdata <- data2[data2$City=='KB',]
jbdata <- data2[data2$City=='JB',]
```

#### 2.1.4 Visualize air pollutant data

```{r}
# Individual plots for air pollutants based on locations

APIKB <- ggplot(kbdata, aes(API))+ geom_histogram(aes(y=..density..)) +
  stat_function(fun=dnorm, colour="magenta", args=list(mean=mean(kbdata$API), sd=sd(kbdata$API))) + labs(x = "API", y = "Case density")

APIJB <- ggplot(jbdata, aes(API))+ geom_histogram(aes(y=..density..)) +
  stat_function(fun=dnorm, colour="magenta", args=list(mean=mean(jbdata$API), sd=sd(jbdata$API))) + labs(x = "API", y = "Case density")

PM10KB <- ggplot(kbdata, aes(PM10))+ geom_histogram(aes(y=..density..)) +
  stat_function(fun=dnorm, colour="magenta", args=list(mean=mean(kbdata$PM10), sd=sd(kbdata$PM10))) + labs(x = "PM10", y = "Case density")

PM10JB <- ggplot(data2[data2$City=='JB',], aes(PM10))+ geom_histogram(aes(y=..density..)) +
  stat_function(fun=dnorm, colour="magenta", args=list(mean=mean(data2[data2$City=='JB',]$PM10), sd=sd(data2[data2$City=='JB',]$PM10))) + labs(x = "PM10", y = "Case density")

PM2.5KB <- ggplot(data2[data2$City=='KB',], aes(PM2.5))+ geom_histogram(aes(y=..density..)) +
  stat_function(fun=dnorm, colour="magenta", args=list(mean=mean(data2[data2$City=='KB',]$PM2.5), sd=sd(data2[data2$City=='KB',]$PM2.5))) + labs(x = "PM2.5", y = "Case density")

PM2.5JB <- ggplot(data2[data2$City=='JB',], aes(PM2.5))+ geom_histogram(aes(y=..density..)) +
  stat_function(fun=dnorm, colour="magenta", args=list(mean=mean(data2[data2$City=='JB',]$PM2.5), sd=sd(data2[data2$City=='JB',]$PM2.5))) + labs(x = "PM2.5", y = "Case density")

SO2KB <- ggplot(data2[data2$City=='KB',], aes(SO2))+ geom_histogram(aes(y=..density..)) +
  stat_function(fun=dnorm, colour="magenta", args=list(mean=mean(data2[data2$City=='KB',]$SO2), sd=sd(data2[data2$City=='KB',]$SO2))) + labs(x = "SO2", y = "Case density")

SO2JB <- ggplot(data2[data2$City=='JB',], aes(SO2))+ geom_histogram(aes(y=..density..)) +
  stat_function(fun=dnorm, colour="magenta", args=list(mean=mean(data2[data2$City=='JB',]$SO2), sd=sd(data2[data2$City=='JB',]$SO2))) + labs(x = "SO2", y = "Case density")

NO2KB <- ggplot(data2[data2$City=='KB',], aes(NO2))+ geom_histogram(aes(y=..density..)) +
  stat_function(fun=dnorm, colour="magenta", args=list(mean=mean(data2[data2$City=='KB',]$NO2), sd=sd(data2[data2$City=='KB',]$NO2))) + labs(x = "NO2", y = "Case density")

NO2JB <- ggplot(data2[data2$City=='JB',], aes(NO2))+ geom_histogram(aes(y=..density..)) +
  stat_function(fun=dnorm, colour="magenta", args=list(mean=mean(data2[data2$City=='JB',]$NO2), sd=sd(data2[data2$City=='JB',]$NO2))) + labs(x = "NO2", y = "Case density")

O3KB <- ggplot(data2[data2$City=='KB',], aes(O3))+ geom_histogram(aes(y=..density..)) +
  stat_function(fun=dnorm, colour="magenta", args=list(mean=mean(data2[data2$City=='KB',]$O3), sd=sd(data2[data2$City=='KB',]$O3))) + labs(x = "O3", y = "Case density")

O3JB <- ggplot(data2[data2$City=='JB',], aes(O3))+ geom_histogram(aes(y=..density..)) +
  stat_function(fun=dnorm, colour="magenta", args=list(mean=mean(data2[data2$City=='JB',]$O3), sd=sd(data2[data2$City=='JB',]$O3))) + labs(x = "O3", y = "Case density")

COKB <- ggplot(data2[data2$City=='KB',], aes(CO3))+ geom_histogram(aes(y=..density..)) +
  stat_function(fun=dnorm, colour="magenta", args=list(mean=mean(data2[data2$City=='KB',]$CO3), sd=sd(data2[data2$City=='KB',]$CO3))) + labs(x = "CO", y = "Case density")

COJB <- ggplot(data2[data2$City=='JB',], aes(CO3))+ geom_histogram(aes(y=..density..)) +
  stat_function(fun=dnorm, colour="magenta", args=list(mean=mean(data2[data2$City=='JB',]$CO3), sd=sd(data2[data2$City=='JB',]$CO))) + labs(x = "CO", y = "Case density")

# Combine plots into one

ggarrange(APIKB, APIJB, PM10KB, PM10JB, PM2.5KB, PM2.5JB, SO2KB, SO2JB, NO2KB, NO2JB, O3KB, O3JB, COKB, COJB,
          heights = 100,
          ncol = 2, nrow = 7)

```

Test for equal variance

```{r}
leveneTest(API ~ Period, data = kbdata)
```

Checking for skewness

```{r}
skewness(data2[data2$City=='JB',]$CO3, na.rm = TRUE)

skewness(data2[data2$City=='KB',]$O3, na.rm = TRUE)
```

Transform skewed data

```{r}
kbdata$O3 <- 1/(kbdata$O3)

jbdata$CO3 <- 1/(jbdata$CO3)
```

Rechecking the data

```{r}
# Distribution of KB O3 variable
ggdensity(kbdata, x = "O3", fill = "lightgray", title = "KB O3") +
  stat_overlay_normal_density(color = "red", linetype = "dashed")

# Distribution of JB CO variable
ggdensity(jbdata, x = "CO3", fill = "lightgray", title = "JB CO") +
  stat_overlay_normal_density(color = "red", linetype = "dashed")

skewness(kbdata$O3, na.rm = TRUE)
skewness(jbdata$CO3, na.rm = TRUE)
```

Time series of air pollutant

```{r}
data$Date <- dmy(data$Date)

APIKB2 <- ggplot(data[data$City=='KB',], aes(x=Date, y=API)) + geom_smooth() + labs(x = "Year") + ylab("Air pollution index")

APIJB2 <- ggplot(data[data$City=='JB',], aes(x=Date, y=API)) + geom_smooth() + labs(x = "Year") + ylab("Air pollution index")

PM10KB2 <- ggplot(data[data$City=='KB',], aes(x=Date, y=PM10)) + geom_smooth()+ labs(x = "Year") + ylab("PM10")

PM10JB2 <- ggplot(data[data$City=='JB',], aes(x=Date, y=PM10)) + geom_smooth() + labs(x = "Year") + ylab("PM10")

PM2.5KB2 <- ggplot(data[data$City=='KB',], aes(x=Date, y=PM2.5)) + geom_smooth() + labs(x = "Year") + ylab("PM2.5")

PM2.5JB2 <- ggplot(data[data$City=='JB',], aes(x=Date, y=PM2.5)) + geom_smooth() + labs(x = "Year") + ylab("PM2.5")

SO2KB2 <- ggplot(data[data$City=='KB',], aes(x=Date, y=SO2)) + geom_smooth() + labs(x = "Year") + ylab("Sulphur dioxide")

SO2JB2 <- ggplot(data[data$City=='JB',], aes(x=Date, y=SO2)) + geom_smooth() + labs(x = "Year") + ylab("Sulphur dioxide")

NO2KB2 <- ggplot(data[data$City=='KB',], aes(x=Date, y=NO2)) + geom_smooth() + labs(x = "Year") + ylab("Nitrogen dioxide")

NO2JB2 <- ggplot(data[data$City=='JB',], aes(x=Date, y=NO2)) + geom_smooth() + labs(x = "Year") + ylab("Nitrogen dioxide")

O3KB2 <- ggplot(data[data$City=='KB',], aes(x=Date, y=O3)) + geom_smooth() + labs(x = "Year") + ylab("Ozone")

O3JB2 <- ggplot(data[data$City=='JB',], aes(x=Date, y=O3)) + geom_smooth() + labs(x = "Year") + ylab("Ozone")

COKB2 <- ggplot(data[data$City=='KB',], aes(x=Date, y=CO3)) + geom_smooth() + labs(x = "Year") + ylab("Carbon monoxide")

COJB2 <- ggplot(data[data$City=='JB',], aes(x=Date, y=CO3)) + geom_smooth() + labs(x = "Year") + ylab("Carbon monoxide")

# Combine plots into one

ggarrange(APIKB2, APIJB2, PM10KB2, PM10JB2, PM2.5KB2, PM2.5JB2, SO2KB2, SO2JB2, NO2KB2, NO2JB2, O3KB2, O3JB2, COKB2, COJB2,
          heights = 100,
          ncol = 2, nrow = 7)
```

#### 2.1.5 Visualize weather data

```{r}

# Individual plots for weather variables based on locations

maxtempkb <- ggplot(data2[data2$City=='KB',], aes(Max))+ geom_histogram(aes(y=..density..)) +
  stat_function(fun=dnorm, colour="magenta", args=list(mean=mean(data2[data2$City=='KB',]$Max), sd=sd(data2[data2$City=='KB',]$Max))) + labs(x = "Max temperature", y = "Case density")

maxtempjb <- ggplot(data2[data2$City=='JB',], aes(Max))+ geom_histogram(aes(y=..density..)) +
  stat_function(fun=dnorm, colour="magenta", args=list(mean=mean(data2[data2$City=='JB',]$Max), sd=sd(data2[data2$City=='JB',]$Max))) + labs(x = "Max temperature", y = "Case density")

meantempkb <- ggplot(data2[data2$City=='KB',], aes(Mean))+ geom_histogram(aes(y=..density..)) +
  stat_function(fun=dnorm, colour="magenta", args=list(mean=mean(data2[data2$City=='KB',]$Mean), sd=sd(data2[data2$City=='KB',]$Mean))) + labs(x = "Mean temperature", y = "Case density")

meantempjb <- ggplot(data2[data2$City=='JB',], aes(Mean))+ geom_histogram(aes(y=..density..)) +
  stat_function(fun=dnorm, colour="magenta", args=list(mean=mean(data2[data2$City=='JB',]$Mean), sd=sd(data2[data2$City=='JB',]$Mean))) + labs(x = "Mean", y = "Case density")

mintempkb <- ggplot(data2[data2$City=='KB',], aes(Mean))+ geom_histogram(aes(y=..density..)) +
  stat_function(fun=dnorm, colour="magenta", args=list(mean=mean(data2[data2$City=='KB',]$Mean), sd=sd(data2[data2$City=='KB',]$Min))) + labs(x = "Min temperature", y = "Case density")

mintempjb <- ggplot(data2[data2$City=='JB',], aes(Min))+ geom_histogram(aes(y=..density..)) +
  stat_function(fun=dnorm, colour="magenta", args=list(mean=mean(data2[data2$City=='JB',]$Min), sd=sd(data2[data2$City=='JB',]$Min))) + labs(x = "Min temperature", y = "Case density")

rainkb <- ggplot(data2[data2$City=='KB',], aes(Rainfall))+ geom_histogram(aes(y=..density..)) +
  stat_function(fun=dnorm, colour="magenta", args=list(mean=mean(data2[data2$City=='KB',]$Rainfall), sd=sd(data2[data2$City=='KB',]$Rainfall))) + labs(x = "Rainfall", y = "Case density")

rainjb <- ggplot(data2[data2$City=='JB',], aes(Rainfall))+ geom_histogram(aes(y=..density..)) +
  stat_function(fun=dnorm, colour="magenta", args=list(mean=mean(data2[data2$City=='JB',]$Rainfall), sd=sd(data2[data2$City=='JB',]$Rainfall))) + labs(x = "Rainfall", y = "Case density")

windkb <- ggplot(data2[data2$City=='KB',], aes(Wind))+ geom_histogram(aes(y=..density..)) +
  stat_function(fun=dnorm, colour="magenta", args=list(mean=mean(data2[data2$City=='KB',]$Wind), sd=sd(data2[data2$City=='KB',]$Wind))) + labs(x = "Wind", y = "Case density")

windjb <- ggplot(data2[data2$City=='JB',], aes(Wind))+ geom_histogram(aes(y=..density..)) +
  stat_function(fun=dnorm, colour="magenta", args=list(mean=mean(data2[data2$City=='JB',]$Wind), sd=sd(data2[data2$City=='JB',]$Wind))) + labs(x = "Wind", y = "Case density")

# Combine plots into one

ggarrange(maxtempkb, maxtempjb, meantempkb, meantempjb, mintempkb, mintempjb, rainkb, rainjb, windkb, windjb,
          heights = 100,
          ncol = 2, nrow = 5)
```

Checking for skewness of rainfall data

```{r}
skewness(kbdata$Rainfall)
skewness(jbdata$Rainfall)
```

Cube root transformation of the skewed data

```{r}
kbdata$Rainfall <- kbdata$Rainfall^(1/3)
jbdata$Rainfall <- jbdata$Rainfall^(1/3)
```

Rechecking for rainfall data

```{r}
# Distribution of KB rainfall
ggdensity(kbdata, x = "Rainfall", fill = "lightgray", title = "Rainfall") +
  stat_overlay_normal_density(color = "red", linetype = "dashed")

# Distribution of JB rainfall
ggdensity(jbdata, x = "Rainfall", fill = "lightgray", title = "Rainfall") +
  stat_overlay_normal_density(color = "red", linetype = "dashed")

skewness(kbdata$Rainfall)
skewness(jbdata$Rainfall)
```

Summary of meteorological variables

```{r}
data2[data2$City=='KB',] %>%
    get_summary_stats(c("Rainfall", "Min", "Max", "Mean", "Rainfall", "Wind"), type = "mean_sd")

data2[data2$City=='JB',] %>%
    get_summary_stats(c("Rainfall", "Min", "Max", "Mean", "Rainfall", "Wind"), type = "mean_sd")
```

Comparison of meteorological variables in JB and KB

```{r}
rainkb <- rnorm(1827, mean = 7.444, sd = 19.74)
rainjb <- rnorm(1827, mean = 7.00, sd = 14.64)
t.test(rainkb, rainjb, var.equal = FALSE)

windkb <- rnorm(1827, mean = 2.31, sd = 0.95)
windjb <- rnorm(1827, mean = 1.53, sd = 0.52)
t.test(windkb, windjb, var.equal = FALSE)

minkb <- rnorm(1827, mean = 24.4, sd = 1.07)
minjb <- rnorm(1827, mean = 23.95, sd = 0.75)
t.test(minkb, minjb, var.equal = FALSE)

meankb <- rnorm(1827, mean = 27.44, sd = 1.17)
meanjb <- rnorm(1827, mean = 27.08, sd = 1.03)
t.test(meankb, meanjb, var.equal = FALSE)

maxkb <- rnorm(1827, mean = 31.4, sd = 1.74)
maxjb <- rnorm(1827, mean = 32.29, sd = 2.8)
t.test(maxkb, maxjb, var.equal = FALSE)
```

#### 2.1.6. Total respiratory diagnosis by city

```{r}
RespiKB <- ggplot(data2[data2$City=='KB',], aes(ED))+ geom_histogram(aes(y=..density..)) +
  stat_function(fun=dnorm, colour="magenta", args=list(mean=mean(data2[data2$City=='KB',]$ED), sd=sd(data2[data2$City=='KB',]$ED))) + labs(x = "Total ED respiratory cases", y = "Case density")


RespiJB <- ggplot(data2[data2$City=='JB',], aes(ED))+ geom_histogram(aes(y=..density..)) +
  stat_function(fun=dnorm, colour="magenta", args=list(mean=mean(data2[data2$City=='JB',]$ED), sd=sd(data2[data2$City=='JB',]$ED))) + labs(x = "Total ED respiratory cases", y = "Case density")


ggarrange(RespiKB, RespiJB,
          heights = 100,
          ncol = 2, nrow = 1)
```

#### 2.1.7. Ranking of respiratory diseases by period

```{r}
my_theme <- function() {

  # Colors
  color.background = "white"
  color.text = "#22211d"

  # Begin construction of chart
  theme_bw(base_size=15) +

    # Format background colors
    theme(panel.background = element_rect(fill=color.background, color=color.background)) +
    theme(plot.background  = element_rect(fill=color.background, color=color.background)) +
    theme(panel.border     = element_rect(color=color.background)) +
    theme(strip.background = element_rect(fill=color.background, color=color.background)) +

    # Format the grid
    theme(panel.grid.major.y = element_blank()) +
    theme(panel.grid.minor.y = element_blank()) +
    theme(axis.ticks       = element_blank()) +

    # Format the legend
    theme(legend.position = "none") +

    # Format title and axis labels
    theme(plot.title       = element_text(color=color.text, size=20, face = "bold")) +
    theme(axis.title.x     = element_text(size=14, color="black", face = "bold")) +
    theme(axis.title.y     = element_text(size=14, color="black", face = "bold", vjust=1.25)) +
    theme(axis.text.x      = element_text(size=10, vjust=0.5, hjust=0.5, color = color.text)) +
    theme(axis.text.y      = element_text(size=10, color = color.text)) +
    theme(strip.text       = element_text(face = "bold")) +

    # Plot margins
    theme(plot.margin = unit(c(0.35, 0.2, 0.3, 0.35), "cm"))
}
```

##### KB rank

```{r}
rankkb <- read.csv('respirankkb.csv')

rankkb$Period2 <- as.factor(rankkb$Period2)

kb.rank <- rankkb %>% 
  group_by(Period) %>% 
  arrange(Period2, Rank, Diagnosis) %>% 
  mutate(ranking = row_number()) %>% 
  as.data.frame()

head(kb.rank)

ggplot(data = kb.rank, aes(x = Period2, y = ranking, group = Diagnosis)) +
  geom_line(aes(color = Diagnosis, alpha = 1), size = 2) +
  geom_point(aes(color = Diagnosis, alpha = 1), size = 3) +
  geom_point(color = "#FFFFFF", size = 1) +
  scale_y_reverse(breaks = 1:8) +
  coord_cartesian(ylim = c(1,8)) + 
  theme(legend.position = "none") +
  my_theme() 

```

##### JB rank

```{r}
rankjb <- read.csv('respirankjb.csv')

rankjb$Period2 <- as.factor(rankjb$Period2)

jb.rank <- rankjb %>% 
  group_by(Period) %>% 
  arrange(Period2, Rank, Diagnosis) %>% 
  mutate(ranking = row_number()) %>% 
  as.data.frame()

head(jb.rank)

ggplot(data = jb.rank, aes(x = Period2, y = ranking, group = Diagnosis)) +
  geom_line(aes(color = Diagnosis, alpha = 1), size = 2) +
  geom_point(aes(color = Diagnosis, alpha = 1), size = 3) +
  geom_point(color = "#FFFFFF", size = 1) +
  scale_y_reverse(breaks = 1:8) +
  coord_cartesian(ylim = c(1,8)) + 
  theme(legend.position = "none") +
  my_theme() 

```

#### 2.1.8. Air pollution by city, by period

```{r}
data2[data2$City=='KB',] %>%
  group_by(Period) %>%
  get_summary_stats(c("API", "PM10", "PM2.5", "SO2", "NO2", "CO3", "O3"), type = "mean_sd")

data2[data2$City=='JB',] %>%
  group_by(Period) %>%
  get_summary_stats(c("API", "PM10", "PM2.5", "SO2", "NO2", "CO3", "O3"), type = "mean_sd")

```

#### 2.1.9. Respiratory diseases by city, by period

```{r}
data2[data2$City=='KB',] %>%
  group_by(Period) %>%
  get_summary_stats(c("Infection", "URTI", "Pneumonia", "LRTI", "CLD", "Symptoms_Signs", "COVID"))

data2[data2$City=='JB',] %>%
  group_by(Period) %>%
  get_summary_stats(c("Infection", "URTI", "Pneumonia", "LRTI", "CLD", "Symptoms_Signs", "COVID"))
```

### 2.2. Variable selection

#### 2.2.1. Correlation

```{r}
# Kota Bharu
##Select numerical variables

datakb <- data2[data2$City=='KB', -1:-12]

##Correlation matrix
cor.datakb <- cor(datakb, use="complete.obs", method="pearson")
head(round(cor.datakb,2))

##Plot correlogram
corrplot(cor.datakb, method="circle")


# Johor Bahru
##Select numerical variables

datajb <- data2[data2$City=='JB', -1:-12]

##Correlation matrix
cor.datajb <- cor(datajb, use="complete.obs", method="pearson")
head(round(cor.datajb,2))

##Plot correlogram
corrplot(cor.datajb, method="circle")
```

Since the air pollution index and particulate matters are highly correlated, two separate analysis with the first containing API only and the other for the separate pollutants will be conducted. These models will then be compared to determine robustness of the models.

Considering that the maximum and minimum temperature are highly correlated with mean temperature, only the mean temperature will be included as covariates in the models.

#### 2.2.2. Comparison of air pollution and respiratory diseases between periods

##### JBKB API t-test

```{r}
statJBAPI <- data2[data2$City=='JB',] %>%
  t_test(API ~ Period) %>%
  add_significance()
statJBAPI

statKBAPI <- data2[data2$City=='KB',] %>%
  t_test(API ~ Period) %>%
  add_significance()
statKBAPI
```

##### JBKB PM10 t-test

```{r}
statJBPM10 <- data2[data2$City=='JB',] %>%
  t_test(PM10 ~ Period) %>%
  add_significance()
statJBPM10

statKBPM10 <- data2[data2$City=='KB',] %>%
  t_test(PM10 ~ Period) %>%
  add_significance()
statKBPM10
```

##### JBKB PM2.5 t-test

```{r}
statJBPM2.5 <- data2[data2$City=='JB',] %>%
  t_test(PM2.5 ~ Period) %>%
  add_significance()
statJBPM2.5

statKBPM2.5 <- data2[data2$City=='KB',] %>%
  t_test(PM2.5 ~ Period) %>%
  add_significance()
statKBPM2.5
```

##### JBKB SO2 t-test

```{r}
statJBSO2 <- data2[data2$City=='JB',] %>%
  t_test(SO2 ~ Period) %>%
  add_significance()
statJBSO2

statKBSO2 <- data2[data2$City=='KB',] %>%
  t_test(SO2 ~ Period) %>%
  add_significance()
statKBSO2
```

##### JBKB NO2 t-test

```{r}
statJBNO2 <- data2[data2$City=='JB',] %>%
  t_test(NO2 ~ Period) %>%
  add_significance()
statJBNO2

statKBNO2 <- data2[data2$City=='KB',] %>%
  t_test(NO2 ~ Period) %>%
  add_significance()
statKBNO2
```

##### JBKB CO t-test

```{r}
statJBCO <- data2[data2$City=='JB',] %>%
  t_test(CO3 ~ Period) %>%
  add_significance()
statJBCO

statKBCO <- data2[data2$City=='KB',] %>%
  t_test(CO3 ~ Period) %>%
  add_significance()
statKBCO
```

##### JBKB 03 t-test

```{r}
statJBO3 <- data2[data2$City=='JB',] %>%
  t_test(O3 ~ Period) %>%
  add_significance()
statJBO3

statKBO3 <- data2[data2$City=='KB',] %>%
  t_test(O3 ~ Period) %>%
  add_significance()
statKBO3
```

##### JBKB infectious or other parasitic t test

```{r}
statJBparasitic <- data2[data2$City=='JB',] %>%
  t_test(Infection ~ Period) %>%
  add_significance()
statJBparasitic

statKBparasitic <- data2[data2$City=='KB',] %>%
  t_test(Infection ~ Period) %>%
  add_significance()
statKBparasitic
```

##### JBKB URTI t test

```{r}
statJBurti <- data2[data2$City=='JB',] %>%
  t_test(URTI ~ Period) %>%
  add_significance()
statJBurti

statKBurti <- data2[data2$City=='KB',] %>%
  t_test(URTI ~ Period) %>%
  add_significance()
statKBurti
```

##### JBKB influenza or pneumonia t test

```{r}
statJBinfluenza <- data2[data2$City=='JB',] %>%
  t_test(Pneumonia ~ Period) %>%
  add_significance()
statJBinfluenza

statKBinfluenza <- data2[data2$City=='KB',] %>%
  t_test(Pneumonia ~ Period) %>%
  add_significance()
statKBinfluenza
```

##### JBKB LRTI t test

```{r}
statJBlrti <- data2[data2$City=='JB',] %>%
  t_test(LRTI ~ Period) %>%
  add_significance()
statJBlrti

statKBlrti <- data2[data2$City=='KB',] %>%
  t_test(LRTI ~ Period) %>%
  add_significance()
statKBlrti
```

##### JBKB CLD t test

```{r}
statJBcld <- data2[data2$City=='JB',] %>%
  t_test(CLD ~ Period) %>%
  add_significance()
statJBcld

statKBcld <- data2[data2$City=='KB',] %>%
  t_test(CLD ~ Period) %>%
  add_significance()
statKBcld
```

##### JBKB Symptoms_Signs t test

```{r}
statJBsx <- data2[data2$City=='JB',] %>%
  t_test(Symptoms_Signs ~ Period) %>%
  add_significance()
statJBsx

statKBsx <- data2[data2$City=='KB',] %>%
  t_test(Symptoms_Signs ~ Period) %>%
  add_significance()
statKBsx
```

##### JBKB covid t test

```{r}
statJBcovid <- data2[data2$City=='JB',] %>%
  t_test(COVID ~ Period) %>%
  add_significance()
statJBcovid

statKBcovid <- data2[data2$City=='KB',] %>%
  t_test(COVID ~ Period) %>%
  add_significance()
statKBcovid
```

##### Multinomial logistic regression for comparing air pollution and respiratory diseases between periods
```{r}
library(VGAM)

respidata <- read.csv("CleanRespi.csv")

respidata$Period <- as.factor(respidata$Period)
respidata$Respiratory.ICD.10 <- as.factor(respidata$Respiratory.ICD.10)
respidata$Hospital <- as.factor(respidata$Hospital)

respidata$Period <- relevel(respidata$Period, ref = "1")
respidata$Respiratory.ICD.10 <- relevel(respidata$Respiratory.ICD.10, ref = "U07.1_COVID19")

kbrespi <- respidata[respidata$Hospital=="HRPZ",]
jbrespi <- respidata[respidata$Hospital=="HSI",]

# Multi-nomial model

modelkb <- vglm(Period ~ RespiDx , multinomial(refLevel = 1), data = kbrespi)
summary(modelkb)

modeljb <- vglm(Period ~ RespiDx , multinomial(refLevel = 1), data = jbrespi)
summary(modeljb)
```

Getting the odds ratio for the models
```{r}
exp(confintvglm(modelkb))
exp(coef(modelkb))

exp(confintvglm(modeljb))
exp(coef(modeljb))

```

#### 2.2.3. Univariate analysis

```{r}
UVKB <- tbl_uvregression(
 data2[data2$City=='KB', c("ED", "API", "PM10", "PM2.5", "SO2", "CO3", "O3", "NO2")],
  method = glm,
  y = ED) %>%
  bold_p() 

UVJB <- tbl_uvregression(
  data2[data2$City=='JB', c("ED", "API", "PM10", "PM2.5", "SO2", "CO3", "O3", "NO2")],
  method = glm,
  y = ED) %>%
  bold_p()

tbl_merge(tbls = list(UVKB, UVJB),
          tab_spanner = c("**Kota Bharu**", "**Johor Bahru**")
          )
```

## 3. Association analysis

Change of variable format

```{r}
kbdata$time <- as.numeric(kbdata$Date)/100
kbdata$doy <- as.numeric(format(kbdata$Date, format = "%j"))
kbdata$year <- format(kbdata$Date, format = "%Y")


jbdata$time <- as.numeric(jbdata$Date)/100
jbdata$doy <- as.numeric(format(jbdata$Date, format = "%j"))
jbdata$year <- format(jbdata$Date, format = "%Y")

kbdata$Rainfall2 <- kbdata$Rainfall^(1/3)
kbdata$O32 <- 1/kbdata$O3
```

### 3.1. KB ED

#### 3.1.1. Model selection and checking

```{r}

kb_total <- gam(ED ~ s(PM10) + s(NO2) + s(CO3) + s(SO2) + s(O3) + s(Mean) + s(Rainfall) + s(Wind) + s(doy, bs="cc", k=365) + s(time, k=80) + Period, data = kbdata, method = "REML", family = 'poisson')
summary(kb_total)

plot.gam(kb_total, shade = TRUE, shade.col = "lightblue", shift = coef(kb_total)[1],  seWithMean = TRUE)

gam.check(kb_total)
```

Checking for concurvity

```{r}
concurvity(kb_total, full = TRUE)

concurvity(kb_total, full = FALSE)
```

A value of more than 0.8 is considered to have high concurvity between 2 variables. 2 pairings are noted to have high concurvity, PM10-PM2.5, and CO-O3. PM2.5 is excluded from the final model. However, Co and O3 is maintained until after comparison with JB data

```{r}

kb_nopm2.5 <- gam(ED ~ s(PM10) + s(NO2) + s(CO3) + s(SO2) + s(O3) + s(Mean) + s(Rainfall) + s(Wind) + s(doy, bs="cc", k=365) + s(time) + Period, data = kbdata, method = "REML", family = 'poisson')
summary(kb_nopm2.5)

plot.gam(kb_nopm2.5, shade = TRUE, shade.col = "lightblue", shift = coef(kb_nopm2.5)[1],  seWithMean = TRUE)

gam.check(kb_nopm2.5)
```

```{r}
concurvity(kb_nopm2.5, full = FALSE)
```
 
 Much better. Proceed with comparison with JB data
 
JB model
```{r}

jb_mod <- gam(ED ~ s(PM10) + s(NO2) + s(CO3) + s(SO2) + s(O3) + s(Mean) + s(Rainfall) + s(Wind) + s(doy, bs="cc", k=365) + s(time, bs ="cc") + Period, data = jbdata, method = "REML", family = 'poisson')
summary(jb_mod)

plot.gam(jb_mod, shade = TRUE, shade.col = "lightblue", shift = coef(jb_mod)[1],  seWithMean = TRUE)

gam.check(jb_mod)

concurvity(jb_mod, full = FALSE)
```

For JB data without PM 2.5, the concurvity analysis shows no pairings of variables with high concurvity, including ozone and carbon monoxide. Therefore, both these variables will be included in the final model for comparison between locations.

Checking for 'Period' as an interaction term
```{r}
kb_mod_period <- gam(ED ~ s(PM10, by=Period) + s(NO2, by=Period) + s(CO3, by=Period) + s(SO2, by=Period) + s(O3, by=Period) + s(Mean) + s(Rainfall) + s(Wind) + s(doy, bs="cc", k=365) + s(time, bs ="bs") + Period, data = kbdata, method = "REML", family = 'poisson')

summary(kb_mod_period)

plot.gam(kb_mod_period, shade = TRUE, shade.col = "lightblue", shift = coef(kb_mod_period)[1],  seWithMean = TRUE, ylim = c(0, 15))

gam.check(kb_mod_period)
```

Checking for the validity of including 'Period' as interaction within the model

```{r}
vis.gam(kb_mod_period, theta = 120, n.grid = 51, lwd = 0.4)
```

Compare models with and without interaction

```{r}
AIC(kb_nopm2.5, kb_mod_period)
```

Since the AIC for the model with interaction is higher compared to without the interaction, the model with interaction will be included as the final model

### JB by period

```{r}
jb_mod_period <- gam(ED ~ s(PM10, by=Period) + s(NO2, by=Period) + s(CO3, by=Period) + s(SO2, by=Period) + s(O3, by=Period) + s(Mean) + s(Rainfall) + s(Wind) + s(doy, bs="cc", k=365) + s(time, bs = "bs") + Period, data = jbdata, method = "REML", family = 'poisson')
summary(jb_mod_period)

plot.gam(jb_mod_period, shade = TRUE, shade.col = "lightblue", shift = coef(jb_mod_period)[1],  seWithMean = TRUE)

gam.check(jb_mod_period)
```

Further analysis based on the different respiratory diagnoses

### 3.2 KB analysis

#### KB URTI by Period

```{r}

kb_URTI_period <- gam(URTI ~ s(PM10, by=Period) + s(NO2, by=Period) + s(CO3, by=Period) + s(SO2, by=Period) + s(O3, by=Period) + s(Mean) + s(Rainfall) + s(Wind) + s(doy, bs="cc", k=365) + s(time, bs ="bs") + Period, data = kbdata, method = "REML", family = 'poisson')
summary(kb_URTI_period)

plot.gam(kb_URTI_period, shade = TRUE, shade.col = "lightblue", shift = coef(kb_URTI_period)[1],  seWithMean = TRUE)

gam.check(kb_URTI_period)
```

#### KB influenza or pneumonia by period

```{r}

kb_pneumo_period <- gam(Pneumonia ~ s(PM10, by=Period) + s(NO2, by=Period) + s(CO3, by=Period) + s(SO2, by=Period) + s(O3, by=Period) + s(Mean) + s(Rainfall) + s(Wind) + s(doy, bs = "cc", k = 365) + s(time, bs="bs") + Period, data = kbdata, method = "REML", family = 'poisson')
summary(kb_pneumo_period)

plot.gam(kb_pneumo_period, shade = TRUE, shade.col = "lightblue", shift = coef(kb_pneumo_period)[1],  seWithMean = TRUE)

gam.check(kb_pneumo_period)
```

#### KB CLD by period

```{r}

kb_CLD_period <- gam(CLD ~ s(PM10, by=Period) + s(NO2, by=Period) + s(CO3, by=Period) + s(SO2, by=Period) + s(O3, by=Period) + s(Mean) + s(Rainfall) + s(Wind) + s(doy, bs="cc", k=365) + s(time, bs="bs") + Period, data = kbdata, method = "REML", family = 'poisson')
summary(kb_CLD_period)

plot.gam(kb_CLD_period, shade = TRUE, shade.col = "lightblue", shift = coef(kb_CLD_period)[1],  seWithMean = TRUE)

gam.check(kb_CLD_period)
```

#### KB LRTI by period

```{r}

kb_LRTI_period <- gam(LRTI ~ s(PM10, by=Period) + s(NO2, by=Period) + s(CO3, by=Period) + s(SO2, by=Period) + s(O3, by=Period) + s(Mean) + s(Rainfall) + s(Wind) + s(doy, bs="cc", k=365) + s(time, bs="bs") + Period, data = kbdata, method = "REML", family = 'poisson')
summary(kb_LRTI_period)

plot.gam(kb_LRTI_period, shade = TRUE, shade.col = "lightblue", shift = coef(kb_LRTI_period)[1],  seWithMean = TRUE)

gam.check(kb_LRTI_period)
```

#### KB SX by period

```{r}

kb_sx_period <- gam(Symptoms_Signs ~ s(PM10, by=Period) + s(NO2, by=Period) + s(CO3, by=Period) + s(SO2, by=Period) + s(O3, by=Period) + s(Mean) + s(Rainfall) + s(Wind) + s(doy, bs="cc", k=365) + s(time, bs="bs") + Period, data = kbdata, method = "REML", family = 'poisson')
summary(kb_sx_period)

plot.gam(kb_sx_period, shade = TRUE, shade.col = "lightblue", shift = coef(kb_sx_period)[1],  seWithMean = TRUE)

gam.check(kb_sx_period)
```

#### JB URTI by Period

```{r}

jb_URTI_period <- gam(URTI ~ s(PM10, by=Period) + s(NO2, by=Period) + s(CO3, by=Period) + s(SO2, by=Period) + s(O3, by=Period) + s(Mean) + s(Rainfall) + s(Wind) + s(doy, bs="cc", k=365) + s(time, bs="bs") + Period, data = jbdata, method = "REML", family = 'poisson')
summary(jb_URTI_period)

plot.gam(jb_URTI_period, shade = TRUE, shade.col = "lightblue", shift = coef(jb_URTI_period)[1],  seWithMean = TRUE)

gam.check(jb_URTI_period)
```

#### KB influenza or pneumonia by period

```{r}

jb_pneumo_period <- gam(Pneumonia ~ s(PM10, by=Period) + s(NO2, by=Period) + s(CO3, by=Period) + s(SO2, by=Period) + s(O3, by=Period) + s(Mean) + s(Rainfall) + s(Wind) + s(doy, bs="cc", k=365) + s(time, bs = "bs") + Period, data = jbdata, method = "REML", family = 'poisson')
summary(jb_pneumo_period)

plot.gam(jb_pneumo_period, shade = TRUE, shade.col = "lightblue", shift = coef(jb_pneumo_period)[1],  seWithMean = TRUE)

gam.check(jb_pneumo_period)
```

### 3.3 Johor Bahru analysis

#### JB CLD by period

```{r}

jb_CLD_period <- gam(CLD ~ s(PM10, by=Period) + s(NO2, by=Period) + s(CO3, by=Period) + s(SO2, by=Period) + s(O3, by=Period) + s(Mean) + s(Rainfall) + s(Wind) + s(doy, bs="cc", k=365) + s(time, bs = "bs") + Period, data = jbdata, method = "REML", family = 'poisson')
summary(jb_CLD_period)

plot.gam(jb_CLD_period, shade = TRUE, shade.col = "lightblue", shift = coef(jb_CLD_period)[1],  seWithMean = TRUE)

gam.check(jb_CLD_period)
```

#### JB LRTI by period

```{r}

jb_LRTI_period <- gam(LRTI ~ s(PM10, by=Period) + s(NO2, by=Period) + s(CO3, by=Period) + s(SO2, by=Period) + s(O3, by=Period) + s(Mean) + s(Rainfall) + s(Wind) + s(doy, bs="cc", k=365) + s(time, bs = "bs") + Period, data = jbdata, method = "REML", family = 'poisson')
summary(jb_LRTI_period)

plot.gam(jb_LRTI_period, shade = TRUE, shade.col = "lightblue", shift = coef(jb_LRTI_period)[1],  seWithMean = TRUE)

gam.check(jb_LRTI_period)
```

#### JB SX by period

```{r}

jb_sx_period <- gam(Symptoms_Signs ~ s(PM10, by=Period) + s(NO2, by=Period) + s(CO3, by=Period) + s(SO2, by=Period) + s(O3, by=Period) + s(Mean) + s(Rainfall) + s(Wind) + s(doy, bs="cc", k=365) + s(time, bs = "bs") + Period, data = jbdata, method = "REML", family = 'poisson')
summary(jb_sx_period)

plot.gam(jb_sx_period, shade = TRUE, shade.col = "lightblue", shift = coef(jb_sx_period)[1],  seWithMean = TRUE)

gam.check(jb_sx_period)
```
